On the entropy of isothermal dusty gas: application to protoplanetary disks

### 2016-08-05 model.tex
minor changes

\subsection{Vertical shear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p s}{\p z} -\frac{\p
    P}{\p z}\frac{\p s}{\p r} \right). 
\end{align}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align*}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
  -\frac{\p\epsilon}{\p z}\frac{\p\ln{\rhog}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r}\left[\frac{\p\epsilon}{\p z} + \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}\right]
  \right\} 
\end{align*}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the second term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient.

@@

\begin{align}\label{vshear_split}
\frac{\left|r\p_z\Omega\right|_{\text{
        dust/gas gradient}}}{\left|r\p_z\Omega\right|_{\text{
        temp. gradient}}} \sim
  \epsilon \frac{\mathrm{max}\left(\delta^2,
    1\right)}{\left|q\right|\left(\epsilon + \delta^2\right)},
\end{align}
where $\delta\equiv \Htilde/\Hgas$. 
Since $|q|=O(1)$ in PPDs, Eq. \ref{vshear_split} indicates that
vertical shear due to variations in the dust-to-gas ratio can be at
most comparable to that due to the radial temperature gradient. 
This only occurs for thin dust layers with $\delta^2\ll \epsilon$ (so 
the above ratio $\sim 1/|q|$). Otherwise, vertical shear is associated
with $\p_rc_s^2$.

\subsection{Vertical buoyancy}
Having identified the entropy of an isothermal dusty gas, we find the
vertical buoyancy frequency $N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p s}{\p z} &=
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
  z}\frac{\p\epsilon}{\p z} \\ &
                                  =
  \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
where the second equality assumes $c_s=c_s(r)$ and the final equality
applies to the equilibria in \S\ref{eqm}. Thus,  
\begin{align*}
N_z\lesssim
O\left(\sqrt{\epsilon_0}\OmK\right). 
\end{align*}
However, for thin dust layers such that $H_\epsilon \ll \Hg$, 
$\mathrm{max}\left(N_z\right)$ may occur outside a finite vertical domain.

%\begin{align}
 % N_z^2 
%\end{align}

>>>>

\subsection{Vertical shear}\label{vertshear}
The mixture possess vertical shear. To see this, we eliminate $\Phi$
between Eq. \ref{steady_momr}---\ref{steady_momz} to 
obtain 
\begin{align}\label{vshear}
  r\frac{\p \Omega^2}{\p z} 
%&= \frac{\p\ln{\rho}}{\p r}\frac{\p}{\p
%    z}\left[c_s^2(1-\tepsilon)\right] - \frac{\p\ln{\rho}}{\p z}
%  \frac{\p}{\p r} \left[c_s^2(1-\tepsilon)\right]\\  
   = \frac{1}{\rho}\left(\frac{\p P}{\p r}\frac{\p s}{\p z} -\frac{\p
    P}{\p z}\frac{\p s}{\p r} \right). 
\end{align}
Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align*}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{
  \frac{\p\epsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
  -\frac{\p\epsilon}{\p z}\frac{\p\ln{\rhog}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\left\{\right\} }
  \left. -\frac{q}{r}\left[\frac{\p\epsilon}{\p z} + \left(1+\epsilon\right)\frac{\p\ln{\rhog}}{\p z}\right]
  \right\} 
\end{align*}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the second term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient.

@@

\begin{align}\label{vshear_split}
  \frac{\left|r\p_z\Omega\right|_{\text{
        dust/gas gradient}}}{\left|r\p_z\Omega\right|_{\text{
        temp. gradient}}} \sim
  \epsilon \frac{\mathrm{max}\left(\delta^2,
    1\right)}{\left|q\right|\left(\epsilon + \delta^2\right)},
\end{align}
where $\delta\equiv \Htilde/\Hgas$. 
Since $|q|=O(1)$ in PPDs, Eq. \ref{vshear_split} indicates that
vertical shear due to variations in the dust-to-gas ratio is at 
most comparable to that due to the radial temperature gradient. 
This only occurs for thin dust layers with $\delta^2\ll \epsilon$ (so
the above ratio $\sim 1/|q|$). Otherwise, vertical shear is 
predominantly associated with  $\p_rc_s^2$.

\subsection{Vertical buoyancy}\label{vbuoyancy}
Having identified the entropy of an isothermal dusty gas, we find the
vertical buoyancy frequency $N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p s}{\p z} &=
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
  z}\frac{\p\epsilon}{\p z} \\ &
                                  =
  \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
where the second equality assumes $c_s=c_s(r)$ and the final equality
applies to the equilibria in \S\ref{eqm}. Thus,  
\begin{align*}
N_z\lesssim
O\left(\sqrt{\epsilon_0}\OmK\right). 
\end{align*}
However, for thin dust layers such that $H_\epsilon \ll Hg$, 
$\mathrm{max}\left(N_z\right)$ may occur outside a finite vertical domain.

 % N_z^2 
%\end{align}

### 2016-08-05 result.tex
minor changes

@@

>>>>

\subsection{Qualitative expectations}
\citetalias{lin15} find the appropriate way to compare 
destabilizing vertical shear to stabilizing vertical buoyancy is
$r\p_z\Omega/\OmK$ against $N_z^2/\OmK^2$. We see from \S\ref{vertshear}
and \S\ref{vbuoyancy} that $|r\p_z\Omega|\sim q h_\mathrm{g}\OmK$ for a
thin, gas dominated disk; while $N_z^2\sim \epsilon\OmK^2$. Thus we
expect dust-induced buoyancy forces to stabilize the disk against the
VSI where $\epsilon \gtrsim h_\mathrm{g}$.

%\begin{align}

@@

\subsection{Validity}
We have considered $\tstop=0$ for simplicity. When $\tstop\neq0$, a 
disk initialized as in \S\ref{eqm} will evolve 
due to dust-gas relative drift. In particular, in the
absence of turbulent/diffusive processes, dust particles settle to the
midplane on a timescale $t_\mathrm{settle}\OmK \sim 1/\tstop\OmK$
\citep{takeuchi02}. For the previous results to be valid (i.e. the
basic state does not evolve significantly) we require VSI growth
timescales $t_\mathrm{grow}\ll t_\mathrm{settle}$. Since the maximum
VSI growth rate is $\sim h_\mathrm{g}\OmK$, we need stopping times
such that 
\begin{align}
  \tstop\OmK \ll h_\mathrm{g}. 
\end{align} 
For thin PPD with $h_\mathrm{g}\sim 0.05$, our dusty VSI results are
valid for $\tstop\OmK < O(10^{-2})$.
